Gusto: a toolkit for compatible finite element dynamical cores

### 2024-02-20 source/paper.tex
a bit of writing

Compatible finite elements provide a route to stable, consistent
discretisations of the partial differential equations governing
geophysical fluids which also have the correct wave propagation speeds
and avoid spurious computational modes, even on non-uniform
grids. These properties, along with conservation of mass and
potentially other quantities such as energy and enstrophy, are
considered essential for the accuracy of the dynamical core of weather
and climate models \citep{staniforth2012horizontal}. Achieving these
properties on a non-uniform grid is important because of parallel
scaling. While the required properties of the discretisation guide the
choice of finite element spaces, there are many options for both the
spatial and temporal discretisation and it is not clear which will be
best for any particular problem. Gusto provides a flexible,
extensible, easy-to-use toolkit of options for rapid prototyping of
different spatial and temporal discretisations based on the compatible
finite element framework applied to geophysical fluid dynamics
equations relevant to numerical weather and climate prediction. In
this paper we introduce Gusto to the community, summarise its current
capabilities and outline our vision for future development.

@@

@@

The finite element discretisation of these equations is formed by
choosing appropriate function spaces for the prognostic variables,
multiplying by test functions from these function spaces and
integrating over the domain. In addition to this, we integrate by
parts when the inter-element continuity of the basis functions is not
sufficient to define the result of applying the differential operator
directly. In line with the theory outlined in section \ref{}, we
choose $\MM{u} \in \mathbb{V}_1$ and $D \in \mathbb{V}_2$ with
$\mathbb{V}_1 \subset H(\text{div})$ and $\mathbb{V}_2 \subset L^2$. The
corresponding test functions are denoted by $\MM{w}$ and $\phi$. This gives

\begin{align}
  \int_\Omega\MM{w}\cdot\pp{\MM{u}}{t}dV + \int_\Omega\MM{w}\cdot(\MM{u}\cdot\nabla)\MM{u} dV + \int_\Omega\MM{w}\cdot f\hat{\MM{k}}\times\MM{u}dV - \int_\Omega \nabla\cdot\MM{w} g(D+b) dV &= 0, \forall \MM{w}\in\mathbb{V}_1 \\
  \int_\Omega\phi\pp{D}{t}dV + \int_\Omega\phi\nabla\cdot(D\MM{u}) dV &= 0, \forall \phi\in\mathbb{V}_2
\end{align}

@@

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
  \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
  \pp{\theta}{t} + \MM{u}\cdot\nabla\theta & = 0, \\
  \Pi^{(1-\kappa)/\kappa} = \frac{R}{p_0}\rho\theta, & 
\end{align}

\subsection{}

\section{Software Design}
\label{sec: design}
Gusto is built on top of the Firedrake package, an automated system
for solving partial differential equations using the finite element
method. This gives many advantages, not least that it provides the
finite element machinery required to translate the discretisations
written in the previous section in to the matrix-vector systems that
we need to solve. This is done using the Unified Form Language (UFL),
a domain specific language that enables the definition of the finite
element forms within Python. The other packages that Firedrake depends
on are summarised in \citet{davies2022towards} ** mention extruded
meshes and PETSc?

\begin{itemize}
\item include schematic
\item generic example - no FML
\item lead in to next section... facilitated by FML
\end{itemize}

>>>>

Compatible finite elements provide a route to stable, consistent
discretisations of the partial differential equations governing
geophysical fluids which also have the correct wave propagation speeds
and avoid spurious computational modes, even on non-uniform
grids. These properties, along with conservation of mass and
potentially other quantities such as energy and enstrophy, are
considered essential for the accuracy of the dynamical core of weather
and climate models \citep{staniforth2012horizontal}. Achieving these
properties on a non-orthogonal grid is important because of parallel
scaling**. While the required properties of the discretisation guide the
choice of finite element spaces, there are many options for both the
spatial and temporal discretisation and it is not clear which will be
best for any particular problem. Gusto provides a flexible,
extensible, easy-to-use toolkit of options for rapid prototyping of
different spatial and temporal discretisations based on the compatible
finite element framework applied to geophysical fluid dynamics
equations relevant to numerical weather and climate prediction. In
this paper we introduce Gusto to the community, summarise its current
capabilities and outline our vision for future development.

@@

The finite element method represents fields on a mesh made of
non-overlapping, conforming cells, or elements. Within each element,
the field is represented as a sum of coefficients multiplied by
polynomial basis functions. The finite element space is specified by
the degree of the polynomial basis functions and the continuity of
these functions between the elements. The partial differential
equations used in weather and climate modelling have multiple
prognostic fields and it is not necessary, or indeed desirable, to use
the same finite element function space for each field. Using different
finite element function spaces for each field gives what is known as a
mixed finite element discretisation. If the finite element function
spaces are chosen so that

@@

@@

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
  \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
  \pp{\theta}{t} + \MM{u}\cdot\nabla\theta & = 0, \\
  \Pi^{(1-\kappa)/\kappa} & = \frac{R}{p_0}\rho\theta, & 
\end{align}
where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure,
typically taken to be *, $R$ is ** and $\kappa$ is **.

\subsection{Finite Element Discretisation}
\label{sec: FEM}

Comparing the equation sets given in the previous sections, we see
several similarities. For example, the prognostic equation for the
velocity always contains a nonlinear transport term, a term due to
rotation and a pressure gradient term. The other prognostic fields
satisfy either an advective or continuity form of the transport
equation. These similarities remain when we formulate the compatible
finite element discretisations, as we shall see below.

The finite element discretisation is formed by choosing appropriate
function spaces for the prognostic variables, multiplying by test
functions from these function spaces and integrating over the
domain. In addition to this, we integrate by parts when the
inter-element continuity of the basis functions is not sufficient to
define the result of applying the differential operator directly.

** Don't do whole equation set - go term by term and try to write as
few equations as possible.

In line with the theory outlined in section \ref{}, we
choose $\MM{u} \in \mathbb{V}_1$ and $D \in \mathbb{V}_2$ with
$\mathbb{V}_1 \subset H(\text{div})$ and $\mathbb{V}_2 \subset L^2$. The
corresponding test functions are denoted by $\MM{w}$ and $\phi$. This gives

\section{Software Design}
\label{sec: design}
Gusto is built on top of the Firedrake package, an automated system
for solving partial differential equations using the finite element
method. This gives many advantages, not least that it provides the
finite element machinery required to translate the discretisations
written in the previous section in to the matrix-vector equations that
we need to solve. This is done using the Unified Form Language (UFL),
a domain specific language that enables the definition of the finite
element forms within Python. The other packages that Firedrake depends
on are summarised in \citet{davies2022towards} but we note that we
make particular use of higher order mesh representation on curved
manifolds (**ref), extruded meshes (**ref) and the PETSc library
solver options (**ref).

Prognostic equation sets are each defined in a python class, making
use of the common finite element forms for the transport and
difffusion terms.

** mention physics source terms?

\begin{itemize}
\item include schematic
\item generic example - no FML
\item lead in to next section... facilitated by FML i.e. timesteppers don't care what equations they're given because forms are labelled
\end{itemize}
